
---
title: "Viz"
author: "Michael Lewis, Aarushi Sahejpal, Katherine Prince"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(sf)
library(terra)
library(dplyr)
library(spData)
library(spDataLarge)
library(tmap)    
library(leaflet) 
library(ggplot2) # tidyverse data visualization package
```

## Virginia

DP03_0018 Estimate!!COMMUTING TO WORK!!Workers 16 years and over
```{r}
library(tidycensus)
options(tigris_use_cache = TRUE)

va_commute <- get_acs(
  geography = "tract", 
  variables = "DP03_0018",
  state = "VA", 
  year = 2021,
  geometry = TRUE
) 


plot(va_commute["estimate"])
```

DP03_0065 Estimate!!INCOME AND BENEFITS (IN 2021 INFLATION ADJUSTED DOLLARS)!!Total households!!With earnings!!Mean earnings (dollars)
```{r}
library(tidycensus)
options(tigris_use_cache = TRUE)

va_income <- get_acs(
  geography = "tract", 
  variables = "DP03_0065",
  state = "VA", 
  year = 2021,
  geometry = TRUE
) 


plot(va_income["estimate"])
```
DP03_0095 Estimate!!HEALTH INSURANCE COVERAGE!!Civilian noninstitutionalized population
```{r}
library(tidycensus)
options(tigris_use_cache = TRUE)

va_insurance <- get_acs(
  geography = "tract", 
  variables = "DP03_0095",
  state = "VA", 
  year = 2021,
  geometry = TRUE
) 


plot(va_insurance["estimate"])
```
DP03_0042P Percent!!INDUSTRY!!Civilian employed population 16 years and over!!Educational services, and health care and social assistance
```{r}
library(tidycensus)
options(tigris_use_cache = TRUE)

va_essential <- get_acs(
  geography = "tract", 
  variables = "DP03_0042P",
  state = "VA", 
  year = 2021,
  geometry = TRUE
) 


plot(va_essential["estimate"])
```
## DC

DP03_0018 Estimate!!COMMUTING TO WORK!!Workers 16 years and over
```{r}
library(tidycensus)
options(tigris_use_cache = TRUE)

dc_commute <- get_acs(
  geography = "tract", 
  variables = "DP03_0018",
  state = "DC", 
  year = 2021,
  geometry = TRUE
) 


plot(dc_commute["estimate"])
```

DP03_0065 Estimate!!INCOME AND BENEFITS (IN 2021 INFLATION ADJUSTED DOLLARS)!!Total households!!With earnings!!Mean earnings (dollars)
```{r}
library(tidycensus)
options(tigris_use_cache = TRUE)

dc_income <- get_acs(
  geography = "tract", 
  variables = "DP03_0065",
  state = "DC", 
  year = 2021,
  geometry = TRUE
) 


plot(dc_income["estimate"])
```

DP03_0095 Estimate!!HEALTH INSURANCE COVERAGE!!Civilian noninstitutionalized population
```{r}
library(tidycensus)
options(tigris_use_cache = TRUE)

dc_insurance <- get_acs(
  geography = "tract", 
  variables = "DP03_0095",
  state = "DC", 
  year = 2021,
  geometry = TRUE
) 


plot(dc_insurance["estimate"])
```

DP03_0042P Percent!!INDUSTRY!!Civilian employed population 16 years and over!!Educational services, and health care and social assistance
```{r}
library(tidycensus)
options(tigris_use_cache = TRUE)

dc_essential <- get_acs(
  geography = "tract", 
  variables = "DP03_0042P",
  state = "DC", 
  year = 2021,
  geometry = TRUE
) 


plot(dc_essential["estimate"])
```

## Maryland

DP03_0018 Estimate!!COMMUTING TO WORK!!Workers 16 years and over
```{r}
library(tidycensus)
options(tigris_use_cache = TRUE)

md_commute <- get_acs(
  geography = "tract", 
  variables = "DP03_0018",
  state = "MD", 
  year = 2021,
  geometry = TRUE
) 


plot(md_commute["estimate"])
```

DP03_0065 Estimate!!INCOME AND BENEFITS (IN 2021 INFLATION ADJUSTED DOLLARS)!!Total households!!With earnings!!Mean earnings (dollars)
```{r}
library(tidycensus)
options(tigris_use_cache = TRUE)

md_income <- get_acs(
  geography = "tract", 
  variables = "DP03_0065",
  state = "MD", 
  year = 2021,
  geometry = TRUE
) 


plot(md_income["estimate"])
```

DP03_0095 Estimate!!HEALTH INSURANCE COVERAGE!!Civilian noninstitutionalized population
```{r}
library(tidycensus)
options(tigris_use_cache = TRUE)

md_insurance <- get_acs(
  geography = "tract", 
  variables = "DP03_0095",
  state = "MD", 
  year = 2021,
  geometry = TRUE
) 


plot(md_insurance["estimate"])
```

DP03_0042P Percent!!INDUSTRY!!Civilian employed population 16 years and over!!Educational services, and health care and social assistance
```{r}
library(tidycensus)
options(tigris_use_cache = TRUE)

md_essential <- get_acs(
  geography = "tract", 
  variables = "DP03_0042P",
  state = "MD", 
  year = 2021,
  geometry = TRUE
) 


plot(md_essential["estimate"])
```
