---
title: "Analyzing the Relationship of Economic Indicators on COVID-19 Outcomes"
author: 
  - Michael Lewis^[SIS/CAS, American University]
  - Your Name2^[American University]
date: "2022-11-28"
abstract: "This is our informative abstract of fewer than 200 words. It describes what we investigate, how we investigate it, and what we find."
output: 
  pdf_document:
    number_sections: true
bibliography: main.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, message=FALSE}
#data import and packages
library(tidyverse)
library(stargazer)
library(AICcmodavg)
library(gridExtra)
read_csv("../data_and_cleaning/covid_econ_data.csv") -> df
```

# Introduction

In this section, we introduce the reader to the phenomenon we investigate. We describe the way in which our analysis contributes to an important intellectual debate, or how it answers a pressing political or social question. We introduce our hypotheses, data, and results. We signpost for the reader what's coming in the rest of the paper.

We remember that our paper is not a mystery novel. We note our core results early and often.

Throughout our paper, we use active, first-person language and avoid the passive voice. For example, we write "we examine the relationship between $X$ and $Y$"; we do not write "the relationship between $X$ and $Y$ was examined." Where we do the analysis, we speak about it transparently.

# [Our Substance and Context Section Title Here]

Here we go deeper into the intellectual debate, the political and social context of our investigation. To give the reader a clear sense of why we are writing this paper, we describe the relevant scholarly, technical, or popular literature. We give this section a meaningful *substantive* title; it is not entitled "Literature Review", for example. We cite at least three published, peer-reviewed scholarly works. For example, we could cite which we discussed in class.[^1] We only cite others' work in our paper when it enhances the reader's understanding of what we, the authors of this paper, are doing. We connect everything we cite to *our* investigation; this is our original research, not a book report or an annotated bibliography.

[^1]: To cite a paper within parentheses, use, e.g., 

In order to integrate citations into the References section below, we add entries into our file `main.bib`. This is a plain-text file that we edit in RStudio. We store `main.bib` in the same folder as our paper's `.Rmd` and `.pdf` files. Its entries are formatted so that they can be knit to `.pdf`; see [https://j.mp/2UzTXEZ](https://www.overleaf.com/learn/latex/Bibliography_management_with_bibtex#The_bibliography_file) for example entries for articles, books, and miscellaneous. We can get these entries automatically from Google Scholar by turning on BibTeX in the Google Scholar Settings - Bibliography Manager. Perhaps we use a tool like free, open-source BibDesk to help us manage the `.bib` file.

# Data and Methods

## Data Sources

This project sources two datasets in order to investigate the relationship between certain economic and social variables American Community Survey (ACS) and COVID-19 data in the DC, Maryland, and Virginia. The American Community Survey (ACS) is acquired from <https://api.census.gov/data/key_signup.html> and COVID-19 Data is acquired from <https://apidocs.covidactnow.org>.

Our final combined data set, with both ACS and COVID-19 data, has 158 total observations with 13 variables of interest.

The American Community Survey is a perpetual survey run by the United States Census Bureau that provides the United States government, and its population, access to information surrounding different economic and social characteristics such as jobs, income, and education on a yearly basis. The ACS contacts over 3.5 million households every year for participation in this survey, where all individuals contacted are legally obligated to answer all the questions in the survey. The sample is selected at random by the Census Bureau, where no address is chosen more than once every 5 years.

Data surrounding the spread and impact of COVID-19 in the United States was sourced from COVID ActNow. This organization is a collection of scientists, pandemic experts, engineers, and public health experts who collect COVID-19 data in the United States, and then subsequently assess its quality. Their works supports federal, state, and local government agencies and their data sets are considered some of the most-trusted regarding the COVID-19 pandemic.

In determining the variables of interest that we wanted to investigate from the American Community Survey, we were most interested in the number of working individuals aged 16 years and older that commute to work, the total mean household earnings, the number of individuals that have health insurance coverage, and the percent of working individuals aged 16 years and older that are employed in an essential worker capacity. The specific variables and their official ACS codes are below:

-   DP03_0018 Estimate, **Commuting to Work**, Workers 16 years and over

-   DP03_0065 Estimate, Income and Benefits (in 2021 inflation adjusted dollars), **Total households, With earnings, Mean earnings** (dollars)

-   DP03_0095 Estimate, **Health Insurance Coverage**, Civilian noninstitutionalized population

-   DP03_0042P Percent, Industry, Civilian **employed** population 16 years and over, **Educational services, and health care and social assistance** (Note: we classify this variable as "essential employees", however, we recognize that this term is imperfect in encapsulating varying definitions of "essential" across three states and 150 plus counties. Nonetheless, we believe that these occupations generally represent what can be considered "essential" occupations during the pandemic)

With the variables selected from the ACS, we first started out with time series COVID-19 data for Maryland, DC, and Virginia. We only selected the year 2021 for COVID-19 data in order to match the ACS data from 2021. We were able to get a daily cumulative case and death count from COVID ActNow, but in order to ensure that our analysis was on the same level as ACS data, we created two new variables -- average cases from 2021 and average deaths from 2021. The specific variables of interest from this data set:

-   Average Cases
-   Average Deaths

For the purpose of building a model to assess the significance of the relationship between multiple economic and social indicators and the spread of COVID in DC, Maryland, and Virginia, we aimed to combine both data sets to prepare it for analysis. We were able to load in both Census ACS data and COVID-19 data from COVID ActNow through their respective APIs, and then subsequently cleaned the source data.

For cleaning the COVID-19 data, we parsed out DC, Maryland, Virginia from our complete time series data from COVID ActNow, then created a separate data set with just these three geographic locations, and computed average cases and deaths for each county during 2021. Then we merged both data sets to combine variables from both sets.

We conduct our analyzing using R version `r paste(R.version$major, R.version$minor, sep = ".")` [@rcoreteam22]. To clean our data we primarily used the `tidyverse` [@wickhamtidyverse] packages as well as `janitor` and `lubridate`. Data collection was aided by the `tidycensus` package, which allowed us to pull direct data from the ACS/US Census Bureau. Data analysis/visualization was conducted using `ggplot2` from the Tidyverse, as well as `stargazer` to create tables [@hlavac18]. Finally, we used the the `keyring`, `httr`, and `jsonlite` packages for API imports from the respective sources.

Since this project aims at investigating the relationship, and its subsequent significance, between certain economic and social indicators and COVID-19 average cases and deaths in different locales within DC, Maryland, and Virginia. Because of this goal, a linear model is the best fit to further analyze our research question.

## Methods

We conduct a Ordinary Least Squares (OLS) regression model to explore the relationship between the aforementioned economic indicators and average COVID outcomes in 2021. We chose this method after examining the raw data and noticing patterns which indicated that a linear model may be appropriate. The below figure illustrates this phenomenon with individuals commuting to work and mean COVID deaths. While there is more variability in cases at the upper-ends, on balance the data follows a linear pattern.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1 <- df %>%
  ggplot(aes(y = average_deaths, x = estimate_number_commuting_to_work)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw() +
  labs(x = "Number of Individuals Commuting to Work", 
       y = "Average COVID Deaths") +
  ggtitle("The Relationship Between COVID Outcomes & Individuals Commuting to Work",
          subtitle = "Data for all counties in DC, MD, VA, in 2021")

p2 <- df %>%
  ggplot(aes(y = average_cases, x = estimate_number_commuting_to_work)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Number of Individuals Commuting to Work", y = "Average COVID Cases") +
  theme_bw()

gridExtra::grid.arrange(p1, p2, ncol = 1)
```

Other variables also show evidence of a linear relationship. In our exploratory data analysis we found a positive association between health insurance coverage (or number of individuals in a given county with health insurance) and COVID deaths and cases. This finding seemed counter intuitive and was something we wanted to examine more closely after controlling for the effect of other predictors.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1 <- df %>%
ggplot(aes(y = average_deaths, x = estimate_health_insurance_coverage)) +
  geom_point() + 
  geom_smooth(method = "lm") + 
  theme_bw() + 
  labs(y = "Average COVID Deaths", x = "Health Insurance Coverage") +
  ggtitle("The Relationship Between COVID Outcomes & Health Insurance Coverage", 
          subtitle = "Data for all counties in DC, MD, VA, in 2021") 

p2 <- df %>%
ggplot(aes(y = average_cases, x = estimate_health_insurance_coverage)) +
  geom_point() + 
  geom_smooth(method = "lm") + 
  theme_bw() + 
  labs(y = "Average COVID Cases", x = "Health Insurance Coverage")  

gridExtra::grid.arrange(p1, p2, ncol = 1)
```

This exploratory data analysis led us to pursue linear regression as a method for analyzing multiple economic indicators and mean COVID outcomes. Specifically, we arrived at three research questions: (1) is health insurance coverage still positively associated with worse COVID outcomes even after controlling for the effects of other predictors, (2) what variables had the largest effect on mean COVID cases and deaths, and (3) how important is wealth in predicting COVID outcomes? The next section will discuss our model and results. 

# Findings

Here, we explain and interpret our results. We try to learn as much as we can about our question as possible, given the data and analysis. We present our results clearly. We interpret them for the reader with precision and circumspection. We avoid making claims that are not substantiated by our data.

Note that this section may be integrated into Section \ref{section:data}, if joining the two improves the overall presentation.

Our results for the `cars` data include estimating the linear model

$$
\begin{aligned}
\widehat{Average \ Cases_i} =  
\beta_0 + \beta_2(Percent \ Essential \ Workers_i) + \beta_2(Number \ of \ Individuals \ with \ Health \ Insurance_i) \\ 
+ \beta_3(Average \ Vaccinations \ Administered_i) + \epsilon_i
\end{aligned}
$$
```{r, message=FALSE, results='asis', echo=FALSE}
average_death.full.lm <-
  lm(
    average_deaths ~ estimate_percent_essential_workers +
      estimate_health_insurance_coverage + estimate_number_commuting_to_work +
      estimate_mean_earnings_usd + average_vaccinations_administered,
    data = df
  )

average_cases.full.lm <-
  lm(
    average_cases ~ estimate_percent_essential_workers +
      estimate_health_insurance_coverage + estimate_number_commuting_to_work +
      estimate_mean_earnings_usd + average_vaccinations_administered,
    data = df
  )

regression_table1 <- stargazer::stargazer(
  average_cases.full.lm,
  average_death.full.lm,
  title = "Two Regression Models Predicting Variation in 2021 COVID Outcomes",
  colnames = FALSE,
  header = FALSE,
  model.numbers = FALSE,
  dep.var.caption = "For all counties in DC, MD, VA",
  dep.var.labels = c("Average Cases", "Average Deaths"),
  covariate.labels = c(
    "Percent Essential Workers",
    "Number of Individuals with Health Insurance",
    "Number of Individuals Commuting to Work",
    "Mean Earnings (USD)",
    "Average Vaccinations Administered",
    "Intercept"
  ),
  keep.stat = c("rsq", "f"),
  notes.align = "l",
  single.row = T,
  add.lines = list(c(
    "Corrected AIC", round(AICc(average_cases.full.lm), 1),  round(AICc(average_death.full.lm), 1)
  ))
)
```


# Discussion

The COVID-19 pandemic has renewed conversations about health equity disparities. As data becomes more abundant, literature has emerged seeking to understand health outcomes, policy interventions, and disparities [@thornton_addressing_2023]. This research project aims to contribute to this growing body of literature by examining the relationship between socio-economic factors and mean COVID outcomes for all counties/wards in Virginia, Maryland, and the District of Columbia. 

Notably, we find strong evidence that suggests that counties with a higher percent of essential workers are predicted to have more cases and deaths on average. The effect, however, is substantially greater for average cases. This finding is consistent with what we would expect, as well as existing theoretical explanations. Indeed, @rosengren_covid-19_2022, @rhodes_occupational_2022, @do_us_2021, and @welsh_workplace_2022 all underscore the elevated risks that essential workers face(d) during the pandemic and outbreaks. Our analysis, however, considers the larger implications of this by estimating how many more cases and deaths a local county could possibly expect based on the percentage of essential workers. While these findings cannot be generalized beyond the mid-Atlantic – and specifically DC, Maryland, and Virginia – it does provide and important contribution in measuring this phenomenon at the local level. 

We also find a puzzling positive association between worse COVID outcomes and health insurance coverage. This finding -- while strange -- is not novel @cuadros_impact_2023 document a similar association between insurance coverage and a higher burden of disease. Importantly, the authors note that there isn’t a “negative impact of insurance on health status” – rather a series of factors which may be at play. First, insured individuals may be more likely to seek care than uninsured individuals – something which is well documented in the consistent and pervasive under-diagnosis of diseases/conditions in the uninsured population. Second, population size may play an intervening role. Larger counties/locales which are predisposed to having more cases could plausibly have a greater raw number of individuals who are insured. Finally, our analysis revealed that health insurance coverage and the number of individuals who commute to work were highly associated. This collinearity makes it difficult to estimate the true effect of health insurance coverage on average COVID outcomes. In short, we don’t place much emphasis on the effects of health insurance (and by extension the number of individuals commuting to work). The literature over overwhelmingly suggests that health insurance plays a positive role in health status @uninsurance_effects_2002, however, statistical results like ours are possible @cuadros_impact_2023 for a variety of reasons. Thus, these results likely indicate there are other factors at play that our model doesn’t adequately capture. 

Our final research question was aimed at understanding the role of earnings on average COVID outcomes. We found that as mean earnings increased counties could expect less cases and deaths on average. On average, an increase of $10,000 (USD) for a county is predicted to result in a decrease of ten deaths in our model. This measure should not be generalized beyond DC, Maryland, and Virginia and should be used with caution. The relationship between wealth and health disparities has been well-documented in both the COVID-19 pandemic and beyond [@tai_disproportionate_2022; @khayat_covid-19_2022]. However, there are likely many variables which we did not include due to project time-constraints, that would enhance our estimate. Specifically, mean earnings vary by location, industry, and population size – all things which future research should consider as factors when estimating the relationship between earnings and health outcomes. Despite this, our estimate contributes to this growing research body and provides a baseline for future analysis. 

Our research aimed at exploring the intersection of economic indicators and average COVID outcomes. To date, much research has been done on the role of socio-economic factors and health outcomes. Our research builds on this by examining this for three states in the mid-Atlantic. Overall, we discovered that the percent of essential workers and mean earnings (at the county level) were important variables when estimating average COVID outcomes. Our models explained between 80-90% of the variation in average COVID outcomes within our sample. Moving forward, future research should be aimed at expanding both the temporal and spatial scopes we employed as well as consider more socio-economic variables which we didn’t have the chance to incorporate. 




# References
